Extracting Angular Observables without a Likelihood\\and Applications to Rare Decays

### 2014-07-17 draft.tex
Update to Sec. III

As before, we assume a PDF $P$ can be decomposed into $N$ parts with angular observables $\lbrace S_k\rbrace$,
\begin{equation}
    P \equiv P(\vec\nu,\vec\theta | \lbrace S_k \rbrace)\,.
\end{equation}
We denote the estimator of an
observable $S_k$ as $\est{S_k}$, which is affected by the detector acceptance. We assume now that there is a linear relation
between the true angular observables $\lbrace S_k^\text{true}\rbrace$, and the estimators $\lbrace \est{S_k}\rbrace$,
\fred{Do we need $S_k^\text{true}$? Isn't it just $S_k$?}
\begin{equation}
    \label{eq:linear-assumption}
    \est{S_i} = \sum_{i = 1}^N M_{i j} S_j^\text{true}\,.
\end{equation}
In that case, the matrix $M_{ij}$ can be fully determined through the following steps.
\begin{enumerate}
    \item For a fixed $j$, let $S_j^\text{true} = 1$, and $S_k^\text{true} = 0\,\forall k\neq j$.
    \item Draw simulated events $\lbrace x_t^{(k)}\rbrace = \lbrace(\vec{\nu}^{(k)}_t,\vec\theta^{(k)}_t)\rbrace$ (dubbed ``true'') from the PDF $P_j$,
        \begin{equation}
            P_j \equiv P(\vec\nu,\vec\theta | \lbrace S_j = 1, S_{k\neq j} = 0\rbrace)\,.
        \end{equation}
    \item Use the true events as input for the MC simulation of the detector, in order to obtain the detector
        events $\lbrace x^{(k)}_d\rbrace = \lbrace(\vec{\nu}^{(k)}_d,\vec\theta^{(k)}_d)\rbrace$.
    \item Use the method of moments to determine the uncorrected observables $\lbrace \est{S_k} \rbrace$, based on the detector
        events $\lbrace x^{(k)}_d\rbrace$.
    \item By virtue of \refeq{linear-assumption} and step 1, the $j$-th column of $M$ has been determined via
        \begin{equation}
            M_{ij} = \est{S_i}\,.
        \end{equation}
        \fred{I wonder if a small sample size also leads to artificial
          nonzero off-diagonal elements of $M$}
    \item Repeat the above steps for all values of $j$ to determine all columns of $M$.
\end{enumerate}

@@

We emphasize that this methods hinges on the assumption in \refeq{linear-assumption}, which might not be fulfilled in general.
In particular, if detector effects are small, of it they can be reduced through other means, our approach is expected to work.
The effects can be considered small if
\begin{equation}
    ||M - \mathbf{1}||_2 \ll N\,.
\end{equation}
Here $||\cdot||_2$ denotes the Frobenius norm.

\danny{$@$Marcin: How well does this work? Is \refeq{linear-assumption} fulfilled in your cases?}

\vspace{\bigskipamount}
\hrule
\danny{Should we remove the remainder of this section?}
\fred{Comment that in this section $\vec\nu$ and $\vec\theta$ are used after ``correction'' has been applied.}
Thanks to accurate detector simulations and control studies, the experimental analyses have become
remarkably good at removing detector acceptance effects. In this section we therefore assume that
there are only a small, residual acceptance effects left that affect the events. In particular, we
assume that the latter can be modeled as a modified probability density,
\begin{equation}
    \label{eq:PDFwithDE}
    P(\nuvec,\thvec) \to P(\nuvec,\thvec) \times \Omega(\thvec),
\end{equation}
where $\Omega$ is independent of the non-angular kinematics $\nuvec$. The latter assumption is only
is invalid in general. However, if residual angular acceptance only weakly depends on $\vec{\nu}$, one
can always choose appropriate bin $\vec{\nu}$ so that $\Omega$ is approximately constant with respect
to $\vec{\nu}$. For such a case we parametrize
\begin{equation}
    \Omega_{m,l}(\thvec) = \prod_{i=1}^{m} \Big[1 + \sum_{j=1}^l \eps_i^{(j)} p_j(\cos\theta_i)\big]
\end{equation}
where $p_j(\cos\theta_i)$ denotes the $j$th Legendre polynomial. Since we only consider residual
acceptance effects, we expect $\eps_i^{(j)} \ll 1$. \footnote{\danny{what are reasonable values/upper limits on the $\eps$?}}\\

We are aware that setting a bin width that fulfills the aforementioned criterion can be at odds with
other requirements. For instance, in exclusive $b\to s\ell^+\ell^-$ decays, the bin width for
$\vec{\nu} = q^2$ needs to cover a sufficiently large part of the phase space beyond the open charm
threshold. We emphasize that the outcome of an analysis as proposed here can easily combine several
bins by means of a weighted average.\\

\paragraph{Uniangular Distribution}
For definiteness and clarity, we consider first an uni-angular distribution $P(\theta)$.
In this case we can parametrize the PDF $P$ as
\begin{equation}
    P(\theta) = S_1 + S_2 \cos\theta + S_2 \cos^2\theta\,,
\end{equation}
which is a parametrization that has been used in the literature for $\bar{B}\to\bar{K}\ell^+\ell^-$ decays \cite{Bobeth:2007dw}.
While this system of angular functions is not orthonormal, it can easily be orthonormalized. For the basis of angular
functions, see \refapp{btokll}.
Considering detector acceptance effects up to second order in $\cos\theta$, the function $\Omega(\vec\theta)$ reads
\begin{equation}
    \Omega_{1,2}(\theta_1) = 1 + \eps_{1}^{(1)} \cos\theta + \frac{\eps_{1}^{(2)}}{2}\big(3 \cos^2\theta - 1\big)\qquad\eps_{1}^{(i)} = \order{\eps}\,.
\end{equation}
Inclusion of $\Omega_{1,2}$ as part of the PDF as proposed in \refeq{PDFwithDE} modifies the MC estimators for the angular observables.
We find
\begin{equation}
\begin{aligned}
    \wwhat{S_1} & = \Big(1 + \frac{\eps_{1}^{(2)}}{2}\Big) \est{S_1} + \frac{9 \eps_{1}^{(2)}}{70} \est{S_3}\\
    \wwhat{S_2} & = \Big(1 - \frac{2 \eps_{1}^{(2)}}{5}\Big) \est{S_2} - \eps_{1}^{(1)} \est{S_1} - \frac{3\eps_{1}^{(1)}}{5} \est{P_3}\\
    \wwhat{S_3} & = \Big(1 - \frac{11\eps_{1}^{(2)}}{14}\Big)\est{S_3} - \frac{3\eps_{1}^{(2)}}{2} \est{S_1} - \eps_{1}^{(1)} \est{P_2}\,.
\end{aligned}
\end{equation}
Here $\lbrace \est{S_i}\rbrace$ are the uncorrected estimators, while $\lbrace \wwhat{S_i}\rbrace$ denote the
acceptance-corrected estimators of the angular observables.\\

\paragraph{Triangular Distribution}

\begin{table*}
    \resizebox{.99\textwidth}{!}{\ensuremath{\input{tab-btokstarll-1to6.tex}}}\\[2.5\medskipamount]
    \resizebox{.99\textwidth}{!}{\ensuremath{\input{tab-btokstarll-7to12.tex}}}\\[2.5\medskipamount]
    \resizebox{.99\textwidth}{!}{\ensuremath{\input{tab-btokstarll-13to18.tex}}}
    \caption{The matrix $N_{ij}$ that appears in \refeq{eff-corrected-est-btokstarll}. Due to spatial
        constraints, we split its display into columns $j=1s,1c,1i,2s,2c,2i$ (top),
        $j=3,4,4i,5,5i,6s$ (middle) and $j=6c,7,7i,8,8i,9$ (bottom).
        \label{tab:eff-corrected-est-btokstarll}
    }
\end{table*}

As an example for a triangular distribution we use the decay $\bar{B}\to\bar{K}^*(\to \bar{K}\pi)\ell^+\ell^-$.
A parametrization of its angular distribution can be found in \refapp{btokstarll}.\\

For this type of distribution, we expand $\Omega_3$ to linear order in $\eps$,
\begin{equation}
    \Omega_{3,2}(\theta_1,\theta_2,\theta_3) = 1 + \sum_{i=1}^3 \sum_{j=1}^2 \eps_i^{(j)} p_j(\cos\theta_i) + \order{\eps^2}\,.
\end{equation}
This leads to a matrix-valued equation that relates the effiency-corrected estimators $\wwhat{S}_i$ to the uncorrected
estimators $\est{S}_j$,
\begin{equation}
    \label{eq:eff-corrected-est-btokstarll}
    \wwhat{S}_i = \sum_{j=1\text{s},\dots\,9} \big(\delta_{ij} + N_{ij}\big) \est{S}_j\,.
\end{equation}
We provide the matrix $N$ in \reftab{eff-corrected-est-btokstarll}.\\
\hrule

>>>>

As before, we assume a PDF $P$ can be decomposed into $N$ parts with angular observables $\lbrace S_k\rbrace$,
\begin{equation}
    P \equiv P(\vec\nu,\vec\theta | \lbrace S_k \rbrace)\,.
\end{equation}
We denote the estimator of an
observable $S_k$ as $\est{S_k}$, which is affected by the detector acceptance. We assume now that there is a linear relation
between the true angular observables $\lbrace S_k^\text{true}\rbrace$, and the estimators $\lbrace \est{S_k}\rbrace$,
\begin{equation}
    \label{eq:linear-assumption}
    \est{S_i} = \sum_{i = 1}^N M_{i j} S_j^\text{true}\,.
\end{equation}
This assumption is based on a smooth angular acceptance function. While this will not be true for the laboratory frame,
smoothness may be recovered in boosted frame, e.g. the rest frame of the decaying particle.
In that case, the matrix $M_{ij}$ can be fully determined through the following steps.
\begin{enumerate}
    \item For a fixed $j$, let $S_j^\text{true} = 1$, and $S_k^\text{true} = 0\,\forall k\neq j$.
    \item Draw simulated events $\lbrace x_t^{(k)}\rbrace = \lbrace(\vec{\nu}^{(k)}_t,\vec\theta^{(k)}_t)\rbrace$ (dubbed ``true'') from the PDF $P_j$,
        \begin{equation}
            P_j \equiv P(\vec\nu,\vec\theta | \lbrace S_j = 1, S_{k\neq j} = 0\rbrace)\,.
        \end{equation}
    \item Use the true events as input for the MC simulation of the detector, in order to obtain the detector
        events $\lbrace x^{(k)}_d\rbrace = \lbrace(\vec{\nu}^{(k)}_d,\vec\theta^{(k)}_d)\rbrace$.
    \item Use the method of moments to determine the uncorrected observables $\lbrace \est{S_k} \rbrace$, based on the detector
        events $\lbrace x^{(k)}_d\rbrace$.
    \item By virtue of \refeq{linear-assumption} and step 1, the $j$-th column of $M$ has been determined via
        \begin{equation}
            M_{ij} = \est{S_i}\,.
        \end{equation}
        \fred{I wonder if a small sample size also leads to artificial
          nonzero off-diagonal elements of $M$}
    \item Repeat the above steps for all values of $j$ to determine all columns of $M$.
\end{enumerate}

@@

We emphasize that this methods hinges on the assumption in \refeq{linear-assumption}, We thus suggest to explicitly
test the assumption. One way of testing involves repeating the previous procedure
with $S_j^\text{true} = 1/x$. If the lineary assumption holds, one should find
\begin{equation}
    M_{ij} = x \est{S_i}\,,
\end{equation}
for arbitrary values of $x > 1$.
